---
title: "Chapter 6"
subtitle: "Introduction to Time Series"
author: "Dr. Joel Fingerman"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

The preceding three chapters of this book dealt with regression methods applied to cross sectional data. When we have cross sectional data our approach to forecasting is to use the
explanatory variables to forecast the value of the dependent variable. This approach to forecasting is often termed _interpolation_.

The other broad method of quantitative business forecasting is to use the history of the data we wish to forecast. We examine the historical pattern of the data we wish to forecast and try
to find a discernible pattern. Using this pattern, we forecast forward the future pattern of the data. This is method of forecasting is often termed _extrapolation_. The use of the historical pattern of the data to be forecast is known as _time series methods_.

\newpage

# Stage 1 Collection and Analysis of Times Series Data

> "The future lies ahead."
>
> `r tufte::quote_footer('--- Mort Sahl')`

   A time series of data is data collected sequentially over equal periods of time. There are many forms of time series: time series can be collected as daily data (such as daily stock market data), weekly (such as weekly receipts data), monthly (such as monthly sales data), quarterly (such as quarterly revenue data), or yearly (such as annual profits data).

   Remember though, the time period of data collection must be consistent. If our time series data are both in monthly and quarterly form then forecasting can be exceedingly
difficult. When you collect time series data be certain that the time period of collection will remain consistent.

   To create a time series of data from our example of Sales Volume data we now collect a series of Sales Volume data from _just one location over consecutive monthly periods_. We collected a data set of four years, 48 consecutive months (48 periods) of Sales Volume.

\newpage

```{r, message=FALSE, error=FALSE, comment=FALSE, warning=FALSE, echo=FALSE}
## Import data from table 6.1 text file
Table_61_path <- path.expand("~/R/forecasting/Table 6.1.txt")
table_61 <- read.delim(Table_61_path, header = FALSE, sep = " ", skip = 1)
 # add column names
colnames(table_61) <- c("Period", "Year 1", "Period", "Year 2", "Period", "Year 3", "Period", "Year 4")

library(knitr)
kable(table_61, caption = "Four Years of Time Series Data")
```


## Graphically displaying Time Series Data: Time Series Plot

The corresponding _time series plot_ or _time series graph_ of the Sales Volume is in Figure 6-1 below.

```{r, message=FALSE, error=FALSE, comment=FALSE, warning=FALSE, echo=FALSE}
## Page 3, create plot
 # First, gather the time series data table into one column.
library(tidyr)
table_61_series <- gather(table_61, Period)
table_61_series <- table_61_series[,-(2:5)]
table_61_series$Period <- as.numeric(rownames(table_61_series))
colnames(table_61_series) <- c("Period", "Sales Volume")

 #Creat chjart
library(ggplot2)
ggplot(data = table_61_series, aes(y = `Sales Volume`, x = Period)) +
        geom_line(size = 1) + 
        ggtitle("Table 6.1: Sales Volume")

```

\newpage

## Tips on Graphing Times Series Data: The Practice

   Good graphical presentations of time series data and forecasts can be very useful
and very powerful in the presentation of ideas. A good layout for a time series graph is
always helpful in the data analysis, model fitting, and forecasting.

   Notice in our graph, that we allowed for additional space to the right of the last
observation of our data series. We have 48 observations but allowed for 54 observations.
Notice too that because the time series graph reveals an upward direction of the data series
we also allowed for additional space above the last observation. The value of the last
observation is 677.4 but allowed for values up to 800. In other words we have provided for
plenty of "white space" for our graph.

   Along the horizontal or time period axis we created horizontal reference lines at
every six periods. It would be too dense to create horizontal reference lines at every of 54
periods. And it would be too sparse to form lines at every 12th period. These are monthly
data so every 6 periods denotes a half year. Period 6 is a June, Period 12 is a December,
Period 18 is June again, and so on.

   Along the vertical or Sales Volume axis we set vertical reference lines at every 100
units. Units of 100 allow for sufficient spacing so the graph does not become crowded, and
yet detailed enough so that we can estimate values with reasonable accuracy.

   In summary, when producing time series graph, try to make the graphs uncluttered
without sacrificing accuracy or information.


## Time Series Data: Time Series Notation



```{marginfigure}
In our example data series,


 $Y_1 = 95.5$
 
 $Y_2 = 127.0$
 
 $Y_3 = 118.0$
 
 $\vdots$
 
 $Y_{48} = 677.4$
```

When discussing observations in our data set we think of the data in the form

$Y_1, Y_2, Y_3, ..., Y_T$
     
        
$Y_1$ is the first observation, $Y_2$ is the second observation, and so on. is the oldest observation
down through $Y_T$ which is the most recent observation.



```{marginfigure}
Past or historical observations

$\dotso$, $Y_{T-1}$, $Y_{T-2}$, $Y_{T-3}$

Most recent observation

$Y_T$

Future observation

$Y_{T+1}$, $Y_{T+2}$, $Y_{T+3}$, $\dotso$
```

The uppercase letter $T$ of $Y_T$ will always mean the most recent observation in our time series. $Y_{T+1}$ means then an observation one period into the future. Thus, $Y_{T+1}$, $Y_{T+2}$, $Y_{T+3}$, denote future
observations one, two, and three periods beyond the present time $Y_T$. Similarly, $Y_{T-1}$ denotes an
observation one period in the past, so that $Y_{T-1}$, $Y_{T-2}$, $Y_{T-3}$, denote observations one, two, and three periods in the past.

We can generalize this notation to:

$Y_{T+\ell}$: A future observation of $\ell$ periods ahead
$Y_{T-\ell}$: A historical observation of $\ell$ periods back

   As we distinguish between $Y_i$ and $\hat{Y_i}$ in regression models, we shall destinguish between $Y_{T+\ell}$ and $\hat{Y_T}{\ell}$ in time series models.
   
$Y_{T+\ell}$ is the future observation of $Y$, $\ell$ period ahead, and unkown at time $T$.

$\hat{Y_T}{\ell}$ is the forecast of the future observation, $\ell$ periods ahead, made at time $T$.


# Stage 1 Continued: The Theory of Collecting & Analyzing Data

   We devote most of this chapter to Stage 1 of the Forecasting Process. With time
series we recommend considerable investigation and analysis, to really understand the time
series data, to really understand what the data are “telling us" before attempting to develop forecasting models. Consequently we shall begin this time series analysis with some easy, straightforward methods of analysis.

## Smothing Out Time Series Data

   Almost all time series data are rough, having a stochastic element to the series. We
shall discuss methods of "smoothing out" the data so that we may observe an underlying
structure or pattern. The first, and simplest, method of smoothing time series data is through averaging.

## Smoothing Out the Data by Averaging

   If we believe there exists some underlying trend or pattern in the time series data,
then by “smoothing the data" we smooth out (or average out) the random variations to reveal
the underlying pattern in the series.


## The Practice: Simple Average Smoothing

   We begin with a time series of 30 observations, $T = 30$, which has a constant
mean, or a slowly changing mean, over time, we call this "horizontal data". 
As in Table 6.2 and Figure 6.3.

```{r, message=FALSE, error=FALSE, comment=FALSE, warning=FALSE, echo=FALSE}
 # Import data from table 6.2 text file
Table_62_path <- path.expand("~/R/forecasting/Table 6.2.txt")
table_62 <- read.delim(Table_62_path, header = FALSE, sep = " ", skip = 1)
 # add column names
colnames(table_62) <- c("Period", "Actual", "Period", "Actual")

library(knitr)
kable(table_62, caption = "Horizontal Data: Table 6.2")
```

```{r, message=FALSE, error=FALSE, comment=FALSE, warning=FALSE, echo=FALSE}
## Page 3, create plot
# First, gather the time series data table into one column.
library(tidyr)
table_62_series <- gather(table_62[,c(2,4)])
table_62_series$key<- as.numeric(rownames(table_62_series))
colnames(table_62_series) <- c("Period", "Horizontal Data")

 #Creat chart
#Creat chart
library(ggplot2)
plot_62 <- ggplot(data = table_62_series, aes(y = `Horizontal Data`, x = Period)) +
        geom_line(size = 1) +
        ggtitle("Figure 6.3: Horizonal Data")

plot_62
```

Becuase the $Y_t$ are varying around a constant mean, a first smoothing of $Y_t$ is just its mean, $\bar{Y}$^[Smoothing using the sample mean
$$\bar{Y} = \sum_{t=1}^{T} Y_{t}$$]


In this example, the sample mean of $\bar{Y} = `r round(mean(table_62_series[,2]), digits = 1)`$.


```{r, message=FALSE, error=FALSE, comment=FALSE, warning=FALSE, echo=FALSE}
ggplot(data = table_62_series, aes(y = `Horizontal Data`, x = Period)) +
        geom_line(size = 1) +
        ggtitle("Horizonal Data with Plotted Mean") +
        geom_hline(aes(yintercept=356.5), color = "red", linetype="dashed", size = 1.3)
```


# Moving Average Smoothings

   The simple average discussed above is a smoothing method over all observations.
A _moving average_ is the technique of creating successive new averages by dropping the
"oldest" observation and adding the most recent observation to calculate the new average.

```{r, message=FALSE, error=FALSE, comment=FALSE, warning=FALSE, echo=FALSE}
## Smoothing moving average, pg 8.
period_3_example <- table_62_series[6:8,]
smooth <- round(mean(table_62_series[6:8,2]), digits = 2)
period_3_example$Smoothing <- c("...",smooth,"...")
colnames(period_3_example) <- c("Period", "Actual", "Smoothing")
```

## 3-period Centered Moving Average Smoothing for period $t$

To begin with an example, examine the notation in the margin^[3-period Centered Moving Average: $$CMA_{t}(3) = \frac{Y_{T+1}+Y_{T}+Y_{T-1}}{3}$$]. Next, we shall use the data from Table 6.2 to illustrate this method. For example, a 3-period Centered Moving Average (at period 7) is:

$CMA_{7}(3) = \frac{Y_{8}+Y_{7}+Y_{6}}{3}$


$CMA_{7}(3) = \frac{364+367+375}{3} = `r smooth`$

```{r, message=FALSE, error=FALSE, comment=FALSE, warning=FALSE, echo=FALSE, fig.margin = TRUE}
kable(period_3_example, align = "l", row.names=FALSE, caption="Horizontal Data Table 6.2 subset: Centered Smoothing Average")
```

   This smoothing method is called a 3-period centered moving average because the computed value is placed at the center, or middle, of the 3 periods being used in the calculations.

The Tufte handout style is a style that Edward Tufte uses in his books and handouts. Tufte's style is known for its extensive use of sidenotes, tight integration of graphics with text, and well-set typography. This style has been implemented in LaTeX and HTML/CSS^[See Github repositories [tufte-latex](https://github.com/tufte-latex/tufte-latex) and [tufte-css](https://github.com/edwardtufte/tufte-css)], respectively. We have ported both implementations into the [**tufte** package](https://github.com/rstudio/tufte). If you want LaTeX/PDF output, you may use the `tufte_handout` format for handouts, and `tufte_book` for books. For HTML output, use `tufte_html`. These formats can be either specified in the YAML metadata at the beginning of an R Markdown document (see an example below), or passed to the `rmarkdown::render()` function. See @R-rmarkdown more information about **rmarkdown**.

```yaml
---
title: "An Example Using the Tufte Style"
author: "John Smith"
output:
  tufte::tufte_handout: default
  tufte::tufte_html: default
---
```

There are two goals of this package:

1. To produce both PDF and HTML output with similar styles from the same R Markdown document;
1. To provide simple syntax to write elements of the Tufte style such as side notes and margin figures, e.g. when you want a margin figure, all you need to do is the chunk option `fig.margin = TRUE`, and we will take care of the deails for you, so you never need to think about `\begin{marginfigure} \end{marginfigure}` or `<span class="marginfigure"> </span>`; the LaTeX and HTML code under the hood may be complicated, but you never need to learn or write such code.

If you have any feature requests or find bugs in **tufte**, please do not hesitate to file them to https://github.com/rstudio/tufte/issues. For general questions, you may ask them on StackOverflow: http://stackoverflow.com/tags/rmarkdown.

# Headings

This style provides first and second-level headings (that is, `#` and `##`), demonstrated in the next section. You may get unexpected output if you try to use `###` and smaller headings.

`r newthought('In his later books')`^[[Beautiful Evidence](http://www.edwardtufte.com/tufte/books_be)], Tufte starts each section with a bit of vertical space, a non-indented paragraph, and sets the first few words of the sentence in small caps. To accomplish this using this style, call the `newthought()` function in **tufte** in an _inline R expression_ `` `r ` `` as demonstrated at the beginning of this paragraph.^[Note you should not assume **tufte** has been attached to your R session. You should either `library(tufte)` in your R Markdown document before you call `newthought()`, or use `tufte::newthought()`.]

# Figures

## Margin Figures

Images and graphics play an integral role in Tufte's work. To place figures in the margin you can use the **knitr** chunk option `fig.margin = TRUE`. For example:

```{r fig-margin, fig.margin = TRUE, fig.cap = "MPG vs horsepower, colored by transmission.", fig.width=3.5, fig.height=3.5, cache=TRUE}
library(ggplot2)
mtcars2 <- mtcars
mtcars2$am <- factor(
  mtcars$am, labels = c('automatic', 'manual')
)
ggplot(mtcars2, aes(hp, mpg, color = am)) +
  geom_point() + geom_smooth() +
  theme(legend.position = 'bottom')
```

Note the use of the `fig.cap` chunk option to provide a figure caption. You can adjust the proportions of figures using the `fig.width` and `fig.height` chunk options. These are specified in inches, and will be automatically scaled down to fit within the handout margin.

## Arbitrary Margin Content

In fact, you can include anything in the margin using the **knitr** engine named `marginfigure`. Unlike R code chunks ```` ```{r} ````, you write a chunk starting with ```` ```{marginfigure} ```` instead, then put the content in the chunk. See an example on the right about the first fundamental theorem of calculus.

```{marginfigure}
We know from _the first fundamental theorem of calculus_ that for $x$ in $[a, b]$:
$$\frac{d}{dx}\left( \int_{a}^{x} f(u)\,du\right)=f(x).$$
```

For the sake of portability between LaTeX and HTML, you should keep the margin content as simple as possible (syntax-wise) in the `marginefigure` blocks. You may use simple Markdown syntax like `**bold**` and `_italic_` text, but please refrain from using footnotes, citations, or block-level elements (e.g. blockquotes and lists) there.

## Full Width Figures

You can arrange for figures to span across the entire page by using the chunk option `fig.fullwidth = TRUE`.

```{r fig-fullwidth, fig.width = 10, fig.height = 2, fig.fullwidth = TRUE, fig.cap = "A full width figure.", warning=FALSE, cache=TRUE}
ggplot(diamonds, aes(carat, price)) + geom_smooth() +
  facet_grid(~ cut)
```

 Other chunk options related to figures can still be used, such as `fig.width`, `fig.cap`, `out.width`, and so on. For full width figures, usually `fig.width` is large and `fig.height` is small. In the above example, the plot size is $10 \times 2$.

## Main Column Figures

Besides margin and full width figures, you can of course also include figures constrained to the main column. This is the default type of figures in the LaTeX/HTML output.

```{r fig-main, fig.cap = "A figure in the main column.", cache=TRUE}
ggplot(diamonds, aes(cut, price)) + geom_boxplot()
```

# Sidenotes

One of the most prominent and distinctive features of this style is the extensive use of sidenotes. There is a wide margin to provide ample room for sidenotes and small figures. Any use of a footnote will automatically be converted to a sidenote. ^[This is a sidenote that was entered using a footnote.] 

If you'd like to place ancillary information in the margin without the sidenote mark (the superscript number), you can use the `margin_note()` function from **tufte** in an inline R expression. `r margin_note("This is a margin note.  Notice that there is no number preceding the note.")` This function does not process the text with Pandoc, so Markdown syntax will not work here. If you need to write anything in Markdown syntax, please use the `marginfigure` block described previously.

# References

References can be displayed as margin notes for HTML output. For example, we can cite R here [@R-base]. To enable this feature, you must set `link-citations: yes` in the YAML metadata, and the version of `pandoc-citeproc` should be at least 0.7.2. You can always install your own version of Pandoc from http://pandoc.org/installing.html if the version is not sufficient. To check the version of `pandoc-citeproc` in your system, you may run this in R:

```{r eval=FALSE}
system2('pandoc-citeproc', '--version')
```

If your version of `pandoc-citeproc` is too low, or you did not set `link-citations: yes` in YAML, references in the HTML output will be placed at the end of the output document.

# Tables

You can use the `kable()` function from the **knitr** package to format tables that integrate well with the rest of the Tufte handout style. The table captions are placed in the margin like figures in the HTML output.

```{r}
knitr::kable(
  mtcars[1:6, 1:6], caption = 'A subset of mtcars.'
)
```

# Block Quotes

We know from the Markdown syntax that paragraphs that start with `>` are converted to block quotes. If you want to add a right-aligned footer for the quote, you may use the function `quote_footer()` from **tufte** in an inline R expression. Here is an example:

> "If it weren't for my lawyer, I'd still be in prison. It went a lot faster with two people digging."
>
> `r tufte::quote_footer('--- Joe Martin')`

Without using `quote_footer()`, it looks like this (the second line is just a normal paragraph):

> "Great people talk about ideas, average people talk about things, and small people talk about wine."
>
> --- Fran Lebowitz

# Responsiveness

The HTML page is responsive in the sense that when the page width is smaller than 760px, sidenotes and margin notes will be hidden by default. For sidenotes, you can click their numbers (the superscripts) to toggle their visibility. For margin notes, you may click the circled plus signs to toggle visibility.

# More Examples

The rest of this document consists of a few test cases to make sure everything still works well in slightly more complicated scenarios. First we generate two plots in one figure environment with the chunk option `fig.show = 'hold'`:

```{r fig-two-together, fig.cap="Two plots in one figure environment.", fig.show='hold', cache=TRUE}
p <- ggplot(mtcars2, aes(hp, mpg, color = am)) +
  geom_point()
p
p + geom_smooth()
```

Then two plots in separate figure environments (the code is identical to the previous code chunk, but the chunk option is the default `fig.show = 'asis'` now):

```{r fig-two-separate, ref.label='fig-two-together', fig.cap=sprintf("Two plots in separate figure environments (the %s plot).", c("first", "second")), cache=TRUE}
```

You may have noticed that the two figures have different captions, and that is because we used a character vector of length 2 for the chunk option `fig.cap` (something like `fig.cap = c('first plot', 'second plot')`).

Next we show multiple plots in margin figures. Similarly, two plots in the same figure environment in the margin:

```{r fig-margin-together, fig.margin=TRUE, fig.show='hold', fig.cap="Two plots in one figure environment in the margin.", fig.width=3.5, fig.height=2.5, cache=TRUE}
p
p + geom_smooth(method = 'lm')
```

Then two plots from the same code chunk placed in different figure environments:

```{r fig-margin-separate, fig.margin=TRUE, fig.cap=sprintf("Two plots in separate figure environments in the margin (the %s plot).", c("first", "second")), fig.width=3.5, fig.height=2.5, cache=TRUE}
knitr::kable(head(iris, 15))
p
knitr::kable(head(iris, 12))
p + geom_smooth(method = 'lm')
knitr::kable(head(iris, 5))
```

We blended some tables in the above code chunk only as _placeholders_ to make sure there is enough vertical space among the margin figures, otherwise they will be stacked tightly together. For a practical document, you should not insert too many margin figures consecutively and make the margin crowded. 

You do not have to assign captions to figures. We show three figures with no captions below in the margin, in the main column, and in full width, respectively.

```{r fig-nocap-margin, fig.margin=TRUE, fig.width=3.5, fig.height=2, cache=TRUE}
# a boxplot of weight vs transmission; this figure
# will be placed in the margin
ggplot(mtcars2, aes(am, wt)) + geom_boxplot() +
  coord_flip()
```
```{r fig-nocap-main, cache=TRUE}
# a figure in the main column
p <- ggplot(mtcars, aes(wt, hp)) + geom_point()
p
```
```{r fig-nocap-fullwidth, fig.fullwidth=TRUE, fig.width=10, fig.height=3, cache=TRUE}
# a fullwidth figure
p + geom_smooth(method = 'lm') + facet_grid(~ gear)
```

# Some Notes on Tufte CSS

There are a few other things in Tufte CSS that we have not mentioned so far. If you prefer `r sans_serif('sans-serif fonts')`, use the function `sans_serif()` in **tufte**. For epigraphs, you may use a pair of underscores to make the paragraph italic in a block quote, e.g.

> _I can win an argument on any topic, against any opponent. People know this, and steer clear of me at parties. Often, as a sign of their great respect, they don't even invite me._
>
> `r quote_footer('--- Dave Barry')`

We hope you will enjoy the simplicity of R Markdown and this R package, and we sincerely thank the authors of the Tufte-CSS and Tufte-LaTeX projects for developing the beautiful CSS and LaTeX classes. Our **tufte** package would not have been possible without their heavy lifting.

To see the R Markdown source of this example document, you may follow [this link to Github](https://github.com/rstudio/tufte/raw/master/inst/rmarkdown/templates/tufte_html/skeleton/skeleton.Rmd), use the wizard in RStudio IDE (`File -> New File -> R Markdown -> From Template`), or open the Rmd file in the package:

```{r eval=FALSE}
file.edit(
  tufte:::template_resources(
    'tufte_html', '..', 'skeleton', 'skeleton.Rmd'
  )
)
```

```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```
